%%%%%%%% Muslims in ML (MusIML) Workshop @ ICML 2026 %%%%%%%
%
% 4-page main + unlimited references. Uses official ICML 2026 style
% (double-blind by default; strips author block unless [accepted] set).
% Style files (icml2026.sty / .bst / algorithm.sty / algorithmic.sty /
% fancyhdr.sty) live alongside this .tex file. Bibliography at
% pivot_ab_references.bib.
%
% Companion submission: paper_musiml_conformal.tex (verifier side).
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{icml2026}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalize,noabbrev]{cleveref}

\icmltitlerunning{Closing Evidence-Blindness in Radiology Generators}

\begin{document}

\twocolumn[
  \icmltitle{Closing Evidence-Blindness in Radiology Report Generators \\
             via Dual-Adversarial Causal-Faithfulness Training}

  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Authors}{anon}
  \end{icmlauthorlist}
  \icmlaffiliation{anon}{Anonymous Institution}
  \icmlcorrespondingauthor{Anonymous Authors}{anon@example.org}
  \icmlkeywords{evidence-blindness, causal-faithfulness, radiology report
                generation, multimodal foundation models, medical AI safety,
                IMG\_correct, counterfactual evaluation, LoRA fine-tuning,
                trustworthy AI, MedGemma}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
Vision--language models that draft radiology reports often produce
text that does not reflect the underlying chest X-ray. Diagnostic
studies have documented this evidence-blindness, and existing fixes
wrap the trained model in inference-time filters or conformal
selection without changing what the model has learned. We intervene
at training time. We present dual-adversarial causal-faithfulness
training, a LoRA fine-tune of MedGemma-4B-IT with two ingredients:
a symmetric adversarial filter that downweights training rows that
either a text-only or an image-only predictor can already solve, and
a causal-faithfulness loss that penalises the model whenever the
probability of the gold token fails to drop under image masking. We
pair the recipe with IMG\_correct, a refinement of the standard
image-masking gap that excludes a known failure mode in which a
consistency-loss-trained model flips its top prediction under masking
without losing support for the right answer. On public-data OpenI
chest radiographs, the method drives IMG\_correct from $2.38\pm0.26$
to $74.96\pm0.83$\,pp at matched token-level accuracy across three
seeds. A cross-counterfactual evaluation reports honestly that the
gain is largely specific to the trained counterfactual: cropping and
centre occlusion collapse it. A retroactive IMG\_correct audit of
$18$ prior verifier cells finds induced-inversion contributions
uniformly small, so the methodological refinement verifies prior
numbers rather than overturning them.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Multimodal medical AI systems produce text that disregards the
visual input: \citet{restrepo2025sms} document this across six
open-source chest-X-ray vision--language models with Selective
Modality Shifting. The phenomenon mirrors the partial-input
shortcuts of NLI \citep{gururangan2018annotation,
poliak2018hypothesis} and the multimodal-reward-model shortcuts of
\citet{li2025sfc}. Two gaps motivate this work. \textbf{First},
existing mitigations are either measurement-only (the diagnostic
literature) or frozen-predictor wrappers (conformal selection and
post-hoc filtering; \citealp{gui2024conformal, li2025conflvlm,
mohri2024conformal}); none modifies what the underlying generator
has learned. \textbf{Second}, the only training-time mitigation at
scale --- the Shortcut-Failure Coefficient (SFC) of
\citet{li2025sfc} --- operates only on the text-only side. We close
both gaps.

\textbf{Contributions.} \textbf{(C1) Dual-adversarial
causal-faithfulness training:} a LoRA fine-tune of MedGemma-4B-IT
\citep{google2024medgemma, hu2022lora} that combines a symmetric
text $+$ image adversarial filter (the image-only branch is the
incremental axis relative to \citealp{li2025sfc}) with a
token-wise correctness-gain margin loss on gold tokens.
\textbf{(C2) IMG\_correct:} a methodological refinement of the
image-masking gap (IMG) that excludes the \emph{induced inversion}
failure mode in which a consistency-loss-trained model flips its
top prediction under masking without losing support for the right
answer. We apply IMG\_correct retroactively to prior verifier
checkpoints and find induced-inversion contributions are uniformly
small (Section~\ref{sec:audit}). We additionally run a
cross-counterfactual evaluation that honestly reports the gain is
largely specific to the trained counterfactual.

\section{Methods}
\label{sec:methods}

We fine-tune MedGemma-4B-IT \citep{google2024medgemma} with LoRA
\citep{hu2022lora} rank $16$ on q/k/v/o projections
($11.9$\,M trainable parameters; full hyperparameters in
the supplementary appendix). The training objective is
\begin{equation}
\mathcal{L}(x) = w(x)\cdot\mathcal{L}_{\mathrm{SFT}}(x) +
\lambda_{\mathrm{faith}}\cdot\mathcal{L}_{\mathrm{faith}}(x),
\end{equation}
with $\lambda_{\mathrm{faith}} = 0.3$.

\paragraph{Dual adversarial filter.} The per-row weight is
$w(x) = \min(w_{\text{text}}(x), w_{\text{image}}(x))$, combining
two shortcut detectors trained independently: a RoBERTa-large
text-only cross-encoder (reusing prior verifier infrastructure
[citation withheld for double-blind review]) and a BiomedCLIP
\citep{zhang2024biomedclip} image-only MLP head. Rows on which
either branch's gold-class probability exceeds $0.7$ are
downweighted to $w_{\min} = 0.2$. On the $20{,}560$-row training
pool, the image-only branch catches $6.5\%$ of rows the text-only
branch misses --- the incremental axis relative to
\citet{li2025sfc} (decomposition in the supplementary appendix).

\paragraph{Causal-faithfulness loss.} The naive consistency loss
$\mathrm{ReLU}(m - \mathrm{KL}(p_{\mathrm{full}}\|p_{\mathrm{alt}}))$
admits a degenerate optimum: the model can satisfy the margin by
being arbitrarily wrong under masking. We replace symmetric KL with
a token-wise correctness-gain margin on gold tokens,
\begin{equation}
\mathcal{L}_{\mathrm{faith}}(x) =
\tfrac{1}{|T|}\!\sum_{t \in T}\mathrm{ReLU}\!\bigl(m -
(\log p_{\mathrm{full}}(y_t) - \log p_{\mathrm{alt}}(y_t))\bigr),
\end{equation}
which couples the divergence requirement to gold-token correctness
on the full input ($m = 0.3$ nats). Image masking uses the
per-channel mean of the batch rather than a literal zero tensor:
post-normalisation zeros are out-of-distribution adversarial input,
whereas the per-channel mean is the model's neutral grey image
(supp.\ appendix).

\paragraph{IMG\_correct.} The standard image-masking gap counts
every $\arg\max$ flip under masking as image-grounded behaviour. A
consistency-loss-trained model can drive image-zeroed accuracy to
zero by always flipping its $\arg\max$ even when its gold-class
probability barely drops. IMG\_correct counts only examples where
(a) the full input is correct, (b) the masked input is incorrect,
\emph{and} (c) the gold-class probability drops by at least margin
$\tau$. We report \emph{induced inversion} as the per-example
contribution caught by IMG but excluded by IMG\_correct (formal
derivation in the supplementary appendix).

\section{Experimental Setup}
\label{sec:setup}

\paragraph{Data.} Training and evaluation use public-data sources
only: OpenI \citep{demner2016openi} ($10{,}839$ rows after
text-target filtering; OpenI test split with $1{,}587$ examples
for in-distribution eval) and ChestX-Det10 silver targets
\citep{liu2020chestxdet} for cross-site eval.

\paragraph{Configurations.} Five training configurations $\times$
three seeds: \texttt{sft\_only} (SFT baseline), \texttt{sft\_dual}
(filter only), \texttt{sft\_faith} (faith loss only),
\texttt{sft\_full} (both; headline configuration), and
\texttt{sft\_full\_sfc} (replaces the discrete dual filter with
continuous SFC weights from \citealp{li2025sfc}).

\paragraph{Metrics.} IMG\_correct at $\tau = 0.1$; standard IMG;
induced inversion; token-level accuracy on gold target tokens.

\section{Results}
\label{sec:results}

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{2.6pt}
\caption{Headline IMG\_correct on the OpenI test split
(token-level, $n = 33{,}173$). Seed-mean $\pm$ standard deviation
over three seeds. Induced-inversion (Inv) contributions are
essentially zero on all faith-loss configurations.}
\label{tab:headline}
\begin{tabular}{lcccc}
\toprule
\textbf{Config} & \textbf{IMG\_corr} & \textbf{IMG} & \textbf{Inv} & \textbf{Acc (\%)}\\
\midrule
sft\_only         & $2.38\pm0.26$  & 0.36  & 0.40 & 74.31 \\
sft\_dual         & $2.10\pm0.24$  & 0.89  & 0.58 & 75.41 \\
\textbf{sft\_faith}    & $\mathbf{74.11\pm0.26}$ & 74.04 & 0.01 & 75.61 \\
\textbf{sft\_full}     & $\mathbf{74.96\pm0.83}$ & 74.96 & 0.01 & 75.48 \\
\textbf{sft\_full\_sfc}& $\mathbf{75.42\pm0.28}$ & 75.43 & 0.01 & 75.44 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Headline (Table~\ref{tab:headline}).} The pre-registered
$\geq 5$\,pp improvement is exceeded by an order of magnitude on
three seeds $\times$ three faith-loss configurations. The full
method \texttt{sft\_full} lifts IMG\_correct by $+72.58$\,pp at
$+1.17$\,pp accuracy over the SFT-only baseline. The
dual-filter-only configuration is statistically indistinguishable
from baseline; the faith loss is the load-bearing component
(\texttt{sft\_faith} alone gives $+71.73$\,pp). Induced inversion
is essentially zero across all faith-loss configurations.

\paragraph{Cross-counterfactual eval (honesty check).} We evaluate
\texttt{sft\_full} (seed $42$) under three counterfactuals the
faith loss did \emph{not} train on. Under \textbf{noise}
replacement IMG\_correct stays at $62.20$\,pp. Under \textbf{crop}
($9\%$ true-image patch retained) it collapses to $4.81$\,pp.
Under \textbf{centre occlusion} ($25\%$ of the image area zeroed)
it collapses to $1.11$\,pp. The trained-mean reference is
$74.00$\,pp on this seed. The gap-to-baseline on crop is
$4.81 - 2.80 = 2.01$\,pp --- an order of magnitude smaller than the
headline gap on the trained counterfactual (cf.\
the supplementary appendix). The model tightly fits the trained
per-channel-mean distribution but does not transfer to localised
counterfactuals.

\paragraph{Cross-site transfer.} On $1{,}122$ ChestX-Det10
silver-target rows, \texttt{sft\_full} attains IMG\_correct
$= 54.62$\,pp at $54.65\%$ accuracy; the matched SFT-only baseline
attains $5.12$\,pp. The cross-site gap of $49.5$\,pp is
$\approx 68\%$ of the in-distribution gap, consistent with partial
transfer (supp.\ appendix).

\paragraph{IMG\_correct retroactive audit.}
\label{sec:audit}
We apply IMG\_correct to $18$ previously reported verifier
configurations (nine v5/v6 headline cells + nine v5 ablations).
Induced-inversion contributions are uniformly $\leq 1.31$\,pp
absolute ($\leq 13.2\%$ of the corresponding IMG). The
pre-registered $\geq 30\%$ hypothesis is refuted on every cell
measured. The \texttt{abl\_no\_consist} cell --- a v5 variant with
the consistency loss removed --- collapses IMG\_correct to
$3.60$\,pp while also reducing induced inversion, the opposite of
what the pre-registered concern would predict (full table in
the supplementary appendix).

\section{Discussion}

\paragraph{What IMG\_correct shows about prior systems.} Prior IMG
numbers on the v5/v6 verifier family were largely measuring real
image-grounded behaviour. The IMG\_correct contribution is
therefore (a) a verification that increases confidence in prior
IMG claims under a strict per-example correctness criterion, and
(b) a default-to-report-alongside-IMG diagnostic that detects the
failure mode if less calibrated training recipes were to introduce
it.

\paragraph{Cross-counterfactual partial generalisation.} The
$+72.58$\,pp headline measures responsiveness to the \emph{trained}
counterfactual, with strong transfer to similar untrained
globally-uniform counterfactuals (noise) and weak transfer to
localised counterfactuals (crop, occlude). The honest read:
training-time counterfactual interventions can tightly fit the
in-distribution counterfactual; a more diverse training mix is the
natural follow-up.

\paragraph{Public-data only.} Reproducibility in medical imaging is
limited by access barriers to MIMIC-CXR-credentialed data
\citep{johnson2019mimiccxr}. Our public-data commitment (OpenI
training $+$ ChestX-Det10 silver cross-site) means any researcher
with comparable cloud compute can re-run the pipeline. This costs
scale (MIMIC-CXR has $\sim 50\times$ more rows), but the
methodology generalises.

\section{Limitations and Conclusion}

\textbf{Limitations.} (L1) The faith loss and IMG\_correct measure
overlapping per-token quantities; the cross-counterfactual eval
quantifies this caveat empirically. (L2) OpenI-only training;
cross-site only on silver targets. (L3) Filter calibration was
trained for our prior verifier task and reused; the
conceptually-aligned per-task filter is an approximation. (L4)
Single backbone; the retroactive audit is on our own verifier
checkpoints only.

We presented dual-adversarial causal-faithfulness training and the
IMG\_correct methodological refinement. Pre-registered empirical
claims, tested transparently: \textbf{(i)} $\geq 5$\,pp
IMG\_correct improvement: \emph{hit by an order of magnitude}.
\textbf{(ii)} $\geq 30\%$ of v5.3's IMG as induced inversion:
\emph{refuted} ($1.1\%$). \textbf{(iii)} the dual filter dominates
the faith loss: \emph{inverted} (faith loss is load-bearing).
IMG\_correct lands independently of the empirical results and is a
reusable audit tool for any IMG-style metric. Counterfactual
evidence-sensitivity testing should precede aggregate-accuracy
reporting for any multimodal medical generator.

\bibliographystyle{icml2026}
\bibliography{pivot_ab_references}

\end{document}
