%%%%%%%% ICML 2026 Workshop Submission (4-page main + unlimited appendix) %%%%%%%
%
% Compatible with AI4Research 2026 (4 pages main + unlimited refs/appendix in
% same PDF) and PhilML 2026 (4 content pages + unlimited refs + supplementary).
% Uses the official ICML 2026 style; double-blind by default.
%
% Style files (icml2026.sty / .bst / algorithm.sty / algorithmic.sty /
% fancyhdr.sty) live alongside this .tex file; bibliography at
% pivot_ab_references.bib.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{icml2026}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalize,noabbrev]{cleveref}

\icmltitlerunning{Supplementary Material}

\begin{document}

\twocolumn[
  \icmltitle{Supplementary Material for: \
             Causally-Faithful Radiology Report Generation via Dual-Adversarial Counterfactual Training}

  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Authors}{anon}
  \end{icmlauthorlist}
  \icmlaffiliation{anon}{Anonymous Institution}
  \icmlcorrespondingauthor{Anonymous Authors}{anon@example.org}
  \icmlkeywords{evidence-blindness, causal-faithfulness, radiology report
                generation, multimodal foundation models, medical AI safety,
                IMG\_correct, counterfactual evaluation, LoRA fine-tuning,
                trustworthy AI, MedGemma}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}


% Supplementary material companion document.

\appendix
\onecolumn

\section{Reproducibility Checklist}
\label{app:repro}

\textbf{Met:} (i) Training data sources are public (OpenI; ChestX-Det10
silver-target for cross-site eval; PadChest-GR is named as deferred
cross-site target; MIMIC-CXR is not used). (ii) Hyperparameters in
Appendix~\ref{app:training}; per-configuration YAML in the released
codebase. (iii) $30$ unit tests pass on CPU. (iv) Modal H100 80\,GB
infrastructure documented; all entrypoints documented.
(v) Per-fine-tune training-step counts, save-every-steps schedule, and
resume-from-checkpoint logic in the released code.

\textbf{Deferred to extended version:} code-release URL, weights-release
URL (anonymized for double-blind review; filled at camera-ready);
cross-site eval on PadChest-GR (no silver targets available in present
infrastructure); three-seed averaging on noise/crop/occlude cross-
counterfactuals (single-seed in \S\ref{sec:results}, three-seed values
in App.~\ref{app:cross-cf}); mech-interp sweep validating a gradient-
norm monitor's AUROC.

\section{Extended Methods}
\label{app:methods}

\subsection{Faith loss derivation and implementation}
\label{app:faith}

The naive consistency loss
$\mathcal{L}_{\mathrm{KL}} = \mathrm{ReLU}(m - \mathrm{KL}(p_{\mathrm{full}} \| p_{\mathrm{alt}}))$
admits a degenerate optimum: a model that always emits a uniform
distribution under masking satisfies any finite KL margin without using
the image. The token-wise correctness-gain margin replaces the
distributional divergence with a per-token signal coupled to gold-token
correctness on the full input. The implementation runs three forward
passes per training step (full, image-zeroed, image-flipped on
laterality rows) and exposes a flipped-row mask that zeroes the flipped
contribution on non-laterality rows. Image-zeroing uses the per-channel
mean of the batch's pixel values, not a literal zero tensor: a literal
zero is, after the processor's normalisation, a specific out-of-
distribution input the model has never seen at training time. The
per-channel mean produces a uniform ``neutral'' image at the model's own
normalisation, which is what we want when we say ``the image is absent.''

\subsection{Dual filter four-cell decomposition}
\label{app:filter}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Four-cell decomposition of which filter branch flags
each training row.} On the $20{,}560$-row training pool, the image-only
branch catches an additional $6.5\%$ of rows that the text-only branch
alone would miss; this is the symmetric-extension contribution axis
relative to \citet{li2025sfc}.}
\label{tab:filter-decomp}
\begin{tabular}{lrr}
\toprule
\textbf{Filter outcome} & \textbf{Count} & \textbf{\% of pool}\\
\midrule
Both filters caught (unimodally solvable on either axis) & $15{,}375$ & $74.8\%$\\
Text-only caught, image-only passed                       & $\phantom{0}2{,}156$  & $10.5\%$\\
\textbf{Image-only caught, text-only passed (novelty axis)} & $\phantom{0}\mathbf{1{,}329}$  & $\mathbf{6.5}\%$\\
Neither caught (genuinely multimodal)                     & $\phantom{0}1{,}700$  & $\phantom{0}8.3\%$\\
\bottomrule
\end{tabular}
\end{table*}

After the subsequent text-target filter restricts training to OpenI rows
with non-empty findings text, the per-row weight distribution within
the surviving $10{,}839$ rows is $95.6\%$ at $w_{\min} = 0.2$ and
$4.4\%$ at $w = 1.0$. The continuous-SFC ablation replaces the discrete
$\{w_{\min}, 1\}$ rule with $\mathrm{SFC}(x) = 1 - p_{\mathrm{text}}^{\mathrm{correct}}(x)$
clipped to $[w_{\min}, 1]$.

\subsection{IMG\_correct derivation and the ``IMG\_correct $>$ IMG'' identity}
\label{app:imgcorrect}

For a binary classifier $f$ with gold class $y$, partition examples into
$\mathrm{cc}/\mathrm{ci}/\mathrm{ic}/\mathrm{ii}$ by whether $f$ is
correct on the full and image-zeroed inputs. Then
$\mathrm{IMG} = (n_{\mathrm{ci}} - n_{\mathrm{ic}})/n \times 100$.
IMG\_correct further partitions $\mathrm{ci}$ into a margin-passing
subset (where $p_{\mathrm{full}}(y) - p_{\mathrm{masked}}(y) \geq \tau$)
and a margin-failing subset (the \emph{induced-inversion} sub-
population):
\begin{align*}
\mathrm{IMG\_correct} &= n_{\mathrm{ci,with\text{-}margin}}/n \times 100,\\
\mathrm{induced\ inv.} &= (n_{\mathrm{ci,no\text{-}margin}} -
                            n_{\mathrm{ci,with\text{-}margin}})/n \times 100.
\end{align*}
\textbf{IMG\_correct can exceed IMG.} When the $\mathrm{ic}$ population
(model becomes correct under masking) is non-trivial and the margin-
failing $\mathrm{ci}$ population is small, IMG\_correct
$> (n_{\mathrm{ci}} - n_{\mathrm{ic}})/n$. This is why several
configurations in the retroactive audit show
IMG\_correct $>$ IMG; it is mathematically valid and does \emph{not}
indicate negative induced inversion. For autoregressive generators we
apply this logic per token on the gold target sequence (with the
causal-LM shift between logits and target IDs).

\subsection{Training hyperparameters and compute}
\label{app:training}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Training hyperparameters across all five
configurations.} The configurations differ only in (i) which weights
JSONL is loaded, (ii) whether the faith loss is enabled, (iii) whether
the flipped counterfactual is enabled, and (iv) the resulting
$\lambda_{\mathrm{faith}}$. All other hyperparameters are identical.}
\label{tab:hyper}
\begin{tabular}{lc}
\toprule
\textbf{Hyperparameter} & \textbf{Value}\\
\midrule
Backbone                          & MedGemma-4B-IT (research-only licence)\\
LoRA rank / alpha                 & $16$ / $32$\\
LoRA targets                      & q\_proj, k\_proj, v\_proj, o\_proj\\
Trainable parameters              & $11.9$\,M ($0.28\%$ of $4.31$\,B base)\\
Optimizer                         & AdamW\\
Learning rate                     & $2\times 10^{-4}$\\
Weight decay                      & $0.0$ on LoRA targets, $0.01$ elsewhere\\
LR schedule                       & cosine to zero, $50$-step linear warm-up\\
Batch size                        & $2$ (grad accum $\times 8$; effective $16$)\\
Max optimizer steps               & $2{,}500$\\
Mixed precision                   & bf16\\
Faith-loss margin (nats)          & $0.3$\\
Image masking                     & per-channel mean of batch pixel values\\
$\lambda_{\mathrm{faith}}$        & $0.3$ when enabled, $0$ otherwise\\
$w_{\min}$ (filter downweight)    & $0.2$\\
Filter activation threshold       & $0.7$ on $p_{\text{branch}}^{\text{correct}}$\\
Random seeds                      & $\{42, 1337, 9001\}$\\
GPU                               & NVIDIA H100 80\,GB (Modal)\\
Wallclock per fine-tune           & $1.5$--$4.0$\,h (faith-loss configs $\sim 3\times$ slower)\\
\bottomrule
\end{tabular}
\end{table*}

Faith-loss configurations are $\sim 3\times$ slower per step than
simple-loss configurations because each step runs three forward passes
(full, image-zeroed, image-flipped). Total Phase 4 training compute:
$\approx \$200$ on Modal H100 80\,GB across $15$ fine-tunes ($5$
configs $\times$ $3$ seeds). Total project envelope (training, primary
evals, retroactive audit, cross-counterfactual, cross-site):
$\approx \$400$.

\section{Full IMG\_correct Retroactive Audit}
\label{app:audit}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Full IMG\_correct retroactive audit, all 18 cells}
(nine verifier configurations $+$ nine v5 ablations, each on $n=2{,}974$
verifier rows). \texttt{Acc full} and \texttt{Acc masked} are accuracies
under the full and image-zeroed inputs, respectively.}
\label{tab:audit-full}
\setlength{\tabcolsep}{4.0pt}
\begin{tabular}{lrrrrrr}
\toprule
\textbf{Config} & \textbf{IMG\_corr} & \textbf{IMG} & \textbf{Inv} & \textbf{Inv/IMG} & \textbf{Acc full} & \textbf{Acc mask}\\
\midrule
\multicolumn{7}{l}{\emph{Headline configurations.}}\\
v5\_0\_base                   & 3.83  & 1.28  & 0.17 & 13.2\% & 92.30 & 91.02\\
v5\_1\_ground                 & 3.40  & 2.62  & 0.13 & \phantom{0}5.1\% & 91.22 & 88.60\\
v5\_2\_real                   & 64.66 & 61.84 & 0.91 & \phantom{0}1.5\% & 92.60 & 30.77\\
\textbf{v5\_3\_contrast}      & \textbf{71.05} & \textbf{68.12} & \textbf{0.74} & \textbf{1.1\%} & 93.01 & 24.88\\
v5\_4\_final                  & 68.43 & 63.55 & 0.40 & \phantom{0}0.6\% & 91.69 & 28.14\\
v6\_0\_3site                  & 74.51 & 70.21 & 0.07 & \phantom{0}0.1\% & 91.06 & 20.85\\
v6\_0\_loo\_no\_openi         & 38.20 & 34.20 & 0.00 & \phantom{0}0.0\% & 87.42 & 53.23\\
v6\_0\_loo\_no\_padchest      & 43.95 & 41.80 & 1.31 & \phantom{0}3.1\% & 92.30 & 50.50\\
v6\_0\_loo\_no\_chestx        & 18.59 & 14.53 & 0.20 & \phantom{0}1.4\% & 81.27 & 66.75\\
\addlinespace
\multicolumn{7}{l}{\emph{v5 ablations.}}\\
abl\_scale\_25                & 69.07 & 65.64 & 0.71 & \phantom{0}1.1\% & 92.70 & 27.07\\
abl\_scale\_50                & 70.51 & 67.59 & 0.81 & \phantom{0}1.2\% & 92.80 & 25.22\\
abl\_scale\_100               & 66.24 & 62.81 & 0.61 & \phantom{0}1.0\% & 92.80 & 29.99\\
abl\_hothresh\_70             & 73.84 & 69.97 & 0.57 & \phantom{0}0.8\% & 92.60 & 22.63\\
\textbf{abl\_hothresh\_80}    & \textbf{77.98} & \textbf{74.78} & 0.74 & \phantom{0}1.0\% & 92.77 & 17.99\\
abl\_no\_contrast             & 63.28 & 60.02 & 0.87 & \phantom{0}1.4\% & 92.77 & 32.75\\
abl\_no\_ground               & 72.53 & 69.64 & 0.50 & \phantom{0}0.7\% & 92.97 & 23.34\\
abl\_no\_uncert               & 73.40 & 70.54 & 0.98 & \phantom{0}1.4\% & 93.17 & 22.63\\
\textbf{abl\_no\_consist}     & \textbf{\phantom{0}3.60}  & \textbf{\phantom{0}2.79}  & 0.30 & \phantom{0}10.8\% & 88.80 & 86.01\\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Distribution of induced-inversion contributions.} Across all
$18$ cells: induced inversion has an absolute range of
$[0.00, 1.31]$\,pp and a relative range of $[0.0\%, 13.2\%]$ of the
corresponding IMG. The cell with the largest relative induced inversion
is \texttt{v5\_0\_base} at $13.2\%$, with the smallest absolute IMG
($1.28$\,pp); fractionally large because any flipping at all under
masking is fractionally large at this scale. The cell with the largest
absolute induced inversion is \texttt{v6\_0\_loo\_no\_padchest} at
$1.31$\,pp ($3.1\%$ of IMG). No cell exceeds the pre-registered
$\geq 30\%$ threshold.

\textbf{The \texttt{abl\_no\_consist} cell is the cleanest
disconfirmation of the pre-registered concern.} Removing the consistency
loss collapses IMG\_correct from $\sim 70$\,pp (nominal v5 family) to
$3.60$\,pp --- near baseline --- while \emph{simultaneously} reducing
induced inversion from $\sim 0.7$\,pp to $0.30$\,pp. The pre-registered
concern was that consistency-loss training \emph{caused} induced
inversion. The empirical pattern is the opposite: with the consistency
loss, IMG\_correct is high and induced inversion is small; without it,
both are small.

\section{Cross-Counterfactual Detail}
\label{app:cross-cf}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Cross-counterfactual eval --- full numbers including
masked accuracy.} Each row uses the same \texttt{sft\_full} seed-42
checkpoint; only the counterfactual transformation differs.}
\label{tab:cross-cf-full}
\setlength{\tabcolsep}{3.0pt}
\begin{tabular}{lcccc}
\toprule
\textbf{Counterfactual} & \textbf{IMG} & \textbf{IMG\_corr} & \textbf{Acc full} & \textbf{Acc mask}\\
\midrule
\textbf{mean (TRAINED)} & 73.97 & 74.00 & 75.48 & \phantom{0}1.51\\
noise                   & 62.10 & 62.20 & 75.48 & 13.38\\
crop                    & \phantom{0}3.64 & \phantom{0}4.81 & 75.48 & 71.83\\
occlude                 & \phantom{0}0.28 & \phantom{0}1.11 & 75.48 & 75.20\\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Baseline anchor.} On the matching crop counterfactual,
\texttt{sft\_only} seed-42 yields IMG $= 0.33$\,pp, IMG\_correct
$= 2.80$\,pp at $73.82\%$ accuracy, induced inversion $= 0.26$\,pp
($n = 105{,}921$ tokens). The cross-counterfactual \texttt{sft\_full}
$-$ \texttt{sft\_only} IMG\_correct gap on crop is
$4.81 - 2.80 = 2.01$\,pp; the gap on the trained-mean counterfactual is
$74.00 - 2.38 = 71.62$\,pp. The order-of-magnitude difference between
trained-mean and crop is the operational meaning of the partial-
circularity finding.

\section{Cross-Site Detail}
\label{app:crosssite}

\textbf{Silver target construction.} ChestX-Det10 test rows in the
standard test set have empty \texttt{evidence\_text} fields. We
constructed \texttt{groundbench\_test\_chestx\_det10\_silver.jsonl}
($1{,}122$ rows) by injecting CheXagent-2-3B
\citep{chen2024chexagent} generations as \texttt{reference\_report};
the silver-target nature is reported explicitly. PadChest-GR cross-site
\citep{castro2024padchestgr} requires a fresh CheXagent generation pass
and is deferred. \textbf{Token-level analysis.} The cross-site $54.65\%$
token-level accuracy is $20$\,pp lower than the in-distribution OpenI
accuracy, reflecting (i) genuine cross-site difficulty (different
camera vendors, different radiologist phrasing) and (ii) the silver-
target nature (the model is being graded against another model's
output, not radiologist gold).

\section{Pre-Registered Hypotheses}
\label{app:prereg}

\textbf{(i) Full method reduces IMG\_correct on OpenI-test by
$\geq 5$\,pp vs.\ SFT-only at matched accuracy.} Status:
\textbf{HIT BY AN ORDER OF MAGNITUDE} on three seeds $\times$ three
faith-loss configurations. Realised: $+72.58$\,pp at $+1.17$\,pp
accuracy.

\textbf{(ii) IMG\_correct refinement reveals $\geq 30\%$ of
\texttt{v5\_3\_contrast}'s IMG as induced inversion.} Status:
\textbf{REFUTED}. Realised: $1.1\%$. Pattern holds across all $18$
audited cells (induced-inversion contributions $\leq 1.31$\,pp
absolute, $\leq 13.2\%$ relative).

\textbf{(iii) Dual filter dominates the faith loss in the ablation.}
Status: \textbf{INVERTED}. The empirical pattern is the opposite:
\texttt{sft\_dual} is statistically indistinguishable from
\texttt{sft\_only}; the faith loss is the load-bearing component
(\texttt{sft\_faith} alone gives $+71.73$\,pp over baseline). The
increment from adding the dual filter on top of the faith loss is
$+0.85$\,pp seed-mean (within one standard deviation of
\texttt{sft\_full}). We retain the dual filter in the recipe because
the four-cell decomposition (App.~\ref{app:filter}) demonstrates a
symmetric-extension contribution axis not addressable by text-only
filtering.

One pre-registration HIT, one REFUTED, one INVERTED --- all reported
transparently. The methodological refinement (IMG\_correct) lands as a
contribution independently of the empirical results.

\end{document}
