%%%%%%%% Muslims in ML (MusIML) Workshop @ ICML 2026 %%%%%%%
%
% 4-page main + unlimited references. Uses official ICML 2026 style
% (double-blind by default; strips author block unless [accepted] set).
% Style files (icml2026.sty / .bst / algorithm.sty / algorithmic.sty /
% fancyhdr.sty) live alongside this .tex file. Bibliography at
% pivot_ab_references.bib.
%
% Companion submission: paper_musiml_conformal.tex (verifier side).
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{icml2026}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalize,noabbrev]{cleveref}

\icmltitlerunning{Closing Evidence-Blindness in Radiology Generators}

\begin{document}

\twocolumn[
  \icmltitle{Closing Evidence-Blindness in Radiology Report Generators \\
             via Dual-Adversarial Causal-Faithfulness Training}

  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Authors}{anon}
  \end{icmlauthorlist}
  \icmlaffiliation{anon}{Anonymous Institution}
  \icmlcorrespondingauthor{Anonymous Authors}{anon@example.org}
  \icmlkeywords{evidence-blindness, causal-faithfulness, radiology report
                generation, multimodal foundation models, medical AI safety,
                IMG\_correct, counterfactual evaluation, LoRA fine-tuning,
                trustworthy AI, MedGemma}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
Vision--language models that draft radiology reports often produce
text that does not reflect the underlying chest X-ray. Diagnostic
studies have documented this evidence-blindness, and existing fixes
wrap the trained model in inference-time filters or conformal
selection without changing what the model has learned. We intervene
at training time. We present dual-adversarial causal-faithfulness
training, a LoRA fine-tune of MedGemma-4B-IT with two ingredients:
a symmetric adversarial filter that downweights training rows that
either a text-only or an image-only predictor can already solve, and
a causal-faithfulness loss that penalises the model whenever the
probability of the gold token fails to drop under image masking. We
pair the recipe with IMG\_correct, a refinement of the standard
image-masking gap that excludes a known failure mode in which a
consistency-loss-trained model flips its top prediction under masking
without losing support for the right answer. On public-data OpenI
chest radiographs, the method drives IMG\_correct from $2.38\pm0.26$
to $74.96\pm0.83$\,pp at matched token-level accuracy across three
seeds. A cross-counterfactual evaluation reports honestly that the
gain is largely specific to the trained counterfactual: cropping and
centre occlusion collapse it. A retroactive IMG\_correct audit of
$18$ prior verifier cells finds induced-inversion contributions
uniformly small, so the methodological refinement verifies prior
numbers rather than overturning them.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Multimodal medical AI systems produce text that disregards the
visual input: \citet{restrepo2025sms} document this across six
open-source chest-X-ray vision--language models with Selective
Modality Shifting. The phenomenon mirrors the partial-input
shortcuts of NLI \citep{gururangan2018annotation,
poliak2018hypothesis} and the multimodal-reward-model shortcuts of
\citet{li2025sfc}. Two gaps motivate this work. \textbf{First},
existing mitigations are either measurement-only (the diagnostic
literature) or frozen-predictor wrappers (conformal selection and
post-hoc filtering; \citealp{gui2024conformal, li2025conflvlm,
mohri2024conformal}); none modifies what the underlying generator
has learned. \textbf{Second}, the only training-time mitigation at
scale --- the Shortcut-Failure Coefficient (SFC) of
\citet{li2025sfc} --- operates only on the text-only side. We close
both gaps.

\textbf{Contributions.} \textbf{(C1) Dual-adversarial
causal-faithfulness training:} a LoRA fine-tune of MedGemma-4B-IT
\citep{google2024medgemma, hu2022lora} that combines a symmetric
text $+$ image adversarial filter (the image-only branch is the
incremental axis relative to \citealp{li2025sfc}) with a
token-wise correctness-gain margin loss on gold tokens.
\textbf{(C2) IMG\_correct:} a methodological refinement of the
image-masking gap (IMG) that excludes the \emph{induced inversion}
failure mode in which a consistency-loss-trained model flips its
top prediction under masking without losing support for the right
answer. We apply IMG\_correct retroactively to prior verifier
checkpoints and find induced-inversion contributions are uniformly
small (Section~\ref{sec:audit}). We additionally run a
cross-counterfactual evaluation that honestly reports the gain is
largely specific to the trained counterfactual.

\section{Methods}
\label{sec:methods}

We fine-tune MedGemma-4B-IT \citep{google2024medgemma} with LoRA
\citep{hu2022lora} rank $16$ on q/k/v/o projections
($11.9$\,M trainable parameters; full hyperparameters in
Appendix~\ref{app:training}). The training objective is
\begin{equation}
\mathcal{L}(x) = w(x)\cdot\mathcal{L}_{\mathrm{SFT}}(x) +
\lambda_{\mathrm{faith}}\cdot\mathcal{L}_{\mathrm{faith}}(x),
\end{equation}
with $\lambda_{\mathrm{faith}} = 0.3$.

\paragraph{Dual adversarial filter.} The per-row weight is
$w(x) = \min(w_{\text{text}}(x), w_{\text{image}}(x))$, combining
two shortcut detectors trained independently: a RoBERTa-large
text-only cross-encoder (reusing prior verifier infrastructure
[citation withheld for double-blind review]) and a BiomedCLIP
\citep{zhang2024biomedclip} image-only MLP head. Rows on which
either branch's gold-class probability exceeds $0.7$ are
downweighted to $w_{\min} = 0.2$. On the $20{,}560$-row training
pool, the image-only branch catches $6.5\%$ of rows the text-only
branch misses --- the incremental axis relative to
\citet{li2025sfc} (decomposition in Appendix~\ref{app:filter}).

\paragraph{Causal-faithfulness loss.} The naive consistency loss
$\mathrm{ReLU}(m - \mathrm{KL}(p_{\mathrm{full}}\|p_{\mathrm{alt}}))$
admits a degenerate optimum: the model can satisfy the margin by
being arbitrarily wrong under masking. We replace symmetric KL with
a token-wise correctness-gain margin on gold tokens,
\begin{equation}
\mathcal{L}_{\mathrm{faith}}(x) =
\tfrac{1}{|T|}\!\sum_{t \in T}\mathrm{ReLU}\!\bigl(m -
(\log p_{\mathrm{full}}(y_t) - \log p_{\mathrm{alt}}(y_t))\bigr),
\end{equation}
which couples the divergence requirement to gold-token correctness
on the full input ($m = 0.3$ nats). Image masking uses the
per-channel mean of the batch rather than a literal zero tensor:
post-normalisation zeros are out-of-distribution adversarial input,
whereas the per-channel mean is the model's neutral grey image
(Appendix~\ref{app:faith}).

\paragraph{IMG\_correct.} The standard image-masking gap counts
every $\arg\max$ flip under masking as image-grounded behaviour. A
consistency-loss-trained model can drive image-zeroed accuracy to
zero by always flipping its $\arg\max$ even when its gold-class
probability barely drops. IMG\_correct counts only examples where
(a) the full input is correct, (b) the masked input is incorrect,
\emph{and} (c) the gold-class probability drops by at least margin
$\tau$. We report \emph{induced inversion} as the per-example
contribution caught by IMG but excluded by IMG\_correct (formal
derivation in Appendix~\ref{app:imgcorrect}).

\section{Experimental Setup}
\label{sec:setup}

\paragraph{Data.} Training and evaluation use public-data sources
only: OpenI \citep{demner2016openi} ($10{,}839$ rows after
text-target filtering; OpenI test split with $1{,}587$ examples
for in-distribution eval) and ChestX-Det10 silver targets
\citep{liu2020chestxdet} for cross-site eval.

\paragraph{Configurations.} Five training configurations $\times$
three seeds: \texttt{sft\_only} (SFT baseline), \texttt{sft\_dual}
(filter only), \texttt{sft\_faith} (faith loss only),
\texttt{sft\_full} (both; headline configuration), and
\texttt{sft\_full\_sfc} (replaces the discrete dual filter with
continuous SFC weights from \citealp{li2025sfc}).

\paragraph{Metrics.} IMG\_correct at $\tau = 0.1$; standard IMG;
induced inversion; token-level accuracy on gold target tokens.

\section{Results}
\label{sec:results}

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{2.6pt}
\caption{Headline IMG\_correct on the OpenI test split
(token-level, $n = 33{,}173$). Seed-mean $\pm$ standard deviation
over three seeds. Induced-inversion (Inv) contributions are
essentially zero on all faith-loss configurations.}
\label{tab:headline}
\begin{tabular}{lcccc}
\toprule
\textbf{Config} & \textbf{IMG\_corr} & \textbf{IMG} & \textbf{Inv} & \textbf{Acc (\%)}\\
\midrule
sft\_only         & $2.38\pm0.26$  & 0.36  & 0.40 & 74.31 \\
sft\_dual         & $2.10\pm0.24$  & 0.89  & 0.58 & 75.41 \\
\textbf{sft\_faith}    & $\mathbf{74.11\pm0.26}$ & 74.04 & 0.01 & 75.61 \\
\textbf{sft\_full}     & $\mathbf{74.96\pm0.83}$ & 74.96 & 0.01 & 75.48 \\
\textbf{sft\_full\_sfc}& $\mathbf{75.42\pm0.28}$ & 75.43 & 0.01 & 75.44 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Headline (Table~\ref{tab:headline}).} The pre-registered
$\geq 5$\,pp improvement is exceeded by an order of magnitude on
three seeds $\times$ three faith-loss configurations. The full
method \texttt{sft\_full} lifts IMG\_correct by $+72.58$\,pp at
$+1.17$\,pp accuracy over the SFT-only baseline. The
dual-filter-only configuration is statistically indistinguishable
from baseline; the faith loss is the load-bearing component
(\texttt{sft\_faith} alone gives $+71.73$\,pp). Induced inversion
is essentially zero across all faith-loss configurations.

\paragraph{Cross-counterfactual eval (honesty check).} We evaluate
\texttt{sft\_full} (seed $42$) under three counterfactuals the
faith loss did \emph{not} train on. Under \textbf{noise}
replacement IMG\_correct stays at $62.20$\,pp. Under \textbf{crop}
($9\%$ true-image patch retained) it collapses to $4.81$\,pp.
Under \textbf{centre occlusion} ($25\%$ of the image area zeroed)
it collapses to $1.11$\,pp. The trained-mean reference is
$74.00$\,pp on this seed. The gap-to-baseline on crop is
$4.81 - 2.80 = 2.01$\,pp --- an order of magnitude smaller than the
headline gap on the trained counterfactual (cf.\
Appendix~\ref{app:cross-cf}). The model tightly fits the trained
per-channel-mean distribution but does not transfer to localised
counterfactuals.

\paragraph{Cross-site transfer.} On $1{,}122$ ChestX-Det10
silver-target rows, \texttt{sft\_full} attains IMG\_correct
$= 54.62$\,pp at $54.65\%$ accuracy; the matched SFT-only baseline
attains $5.12$\,pp. The cross-site gap of $49.5$\,pp is
$\approx 68\%$ of the in-distribution gap, consistent with partial
transfer (supp.\ appendix).

\paragraph{IMG\_correct retroactive audit.}
\label{sec:audit}
We apply IMG\_correct to $18$ previously reported verifier
configurations (nine v5/v6 headline cells + nine v5 ablations).
Induced-inversion contributions are uniformly $\leq 1.31$\,pp
absolute ($\leq 13.2\%$ of the corresponding IMG). The
pre-registered $\geq 30\%$ hypothesis is refuted on every cell
measured. The \texttt{abl\_no\_consist} cell --- a v5 variant with
the consistency loss removed --- collapses IMG\_correct to
$3.60$\,pp while also reducing induced inversion, the opposite of
what the pre-registered concern would predict (full table in
Appendix~\ref{app:audit}).

\section{Discussion}

\paragraph{What IMG\_correct shows about prior systems.} Prior IMG
numbers on the v5/v6 verifier family were largely measuring real
image-grounded behaviour. The IMG\_correct contribution is
therefore (a) a verification that increases confidence in prior
IMG claims under a strict per-example correctness criterion, and
(b) a default-to-report-alongside-IMG diagnostic that detects the
failure mode if less calibrated training recipes were to introduce
it.

\paragraph{Cross-counterfactual partial generalisation.} The
$+72.58$\,pp headline measures responsiveness to the \emph{trained}
counterfactual, with strong transfer to similar untrained
globally-uniform counterfactuals (noise) and weak transfer to
localised counterfactuals (crop, occlude). The honest read:
training-time counterfactual interventions can tightly fit the
in-distribution counterfactual; a more diverse training mix is the
natural follow-up.

\paragraph{Public-data only.} Reproducibility in medical imaging is
limited by access barriers to MIMIC-CXR-credentialed data
\citep{johnson2019mimiccxr}. Our public-data commitment (OpenI
training $+$ ChestX-Det10 silver cross-site) means any researcher
with comparable cloud compute can re-run the pipeline. This costs
scale (MIMIC-CXR has $\sim 50\times$ more rows), but the
methodology generalises.

\section{Limitations and Conclusion}

\textbf{Limitations.} (L1) The faith loss and IMG\_correct measure
overlapping per-token quantities; the cross-counterfactual eval
quantifies this caveat empirically. (L2) OpenI-only training;
cross-site only on silver targets. (L3) Filter calibration was
trained for our prior verifier task and reused; the
conceptually-aligned per-task filter is an approximation. (L4)
Single backbone; the retroactive audit is on our own verifier
checkpoints only.

We presented dual-adversarial causal-faithfulness training and the
IMG\_correct methodological refinement. Pre-registered empirical
claims, tested transparently: \textbf{(i)} $\geq 5$\,pp
IMG\_correct improvement: \emph{hit by an order of magnitude}.
\textbf{(ii)} $\geq 30\%$ of v5.3's IMG as induced inversion:
\emph{refuted} ($1.1\%$). \textbf{(iii)} the dual filter dominates
the faith loss: \emph{inverted} (faith loss is load-bearing).
IMG\_correct lands independently of the empirical results and is a
reusable audit tool for any IMG-style metric. Counterfactual
evidence-sensitivity testing should precede aggregate-accuracy
reporting for any multimodal medical generator.

\appendix
\onecolumn

\section{Reproducibility Checklist}
\label{app:repro}

\textbf{Met:} (i) Training data sources are public (OpenI; ChestX-Det10
silver-target for cross-site eval; PadChest-GR is named as deferred
cross-site target; MIMIC-CXR is not used). (ii) Hyperparameters in
Appendix~\ref{app:training}; per-configuration YAML in the released
codebase. (iii) $30$ unit tests pass on CPU. (iv) Modal H100 80\,GB
infrastructure documented; all entrypoints documented.
(v) Per-fine-tune training-step counts, save-every-steps schedule, and
resume-from-checkpoint logic in the released code.

\textbf{Deferred to extended version:} code-release URL, weights-release
URL (anonymized for double-blind review; filled at camera-ready);
cross-site eval on PadChest-GR (no silver targets available in present
infrastructure); three-seed averaging on noise/crop/occlude cross-
counterfactuals (single-seed in the main paper's Results section, three-seed values
in App.~\ref{app:cross-cf}); mech-interp sweep validating a gradient-
norm monitor's AUROC.

\section{Extended Methods}
\label{app:methods}

\subsection{Faith loss derivation and implementation}
\label{app:faith}

The naive consistency loss
$\mathcal{L}_{\mathrm{KL}} = \mathrm{ReLU}(m - \mathrm{KL}(p_{\mathrm{full}} \| p_{\mathrm{alt}}))$
admits a degenerate optimum: a model that always emits a uniform
distribution under masking satisfies any finite KL margin without using
the image. The token-wise correctness-gain margin replaces the
distributional divergence with a per-token signal coupled to gold-token
correctness on the full input. The implementation runs three forward
passes per training step (full, image-zeroed, image-flipped on
laterality rows) and exposes a flipped-row mask that zeroes the flipped
contribution on non-laterality rows. Image-zeroing uses the per-channel
mean of the batch's pixel values, not a literal zero tensor: a literal
zero is, after the processor's normalisation, a specific out-of-
distribution input the model has never seen at training time. The
per-channel mean produces a uniform ``neutral'' image at the model's own
normalisation, which is what we want when we say ``the image is absent.''

\subsection{Dual filter four-cell decomposition}
\label{app:filter}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Four-cell decomposition of which filter branch flags
each training row.} On the $20{,}560$-row training pool, the image-only
branch catches an additional $6.5\%$ of rows that the text-only branch
alone would miss; this is the symmetric-extension contribution axis
relative to \citet{li2025sfc}.}
\label{tab:filter-decomp}
\begin{tabular}{lrr}
\toprule
\textbf{Filter outcome} & \textbf{Count} & \textbf{\% of pool}\\
\midrule
Both filters caught (unimodally solvable on either axis) & $15{,}375$ & $74.8\%$\\
Text-only caught, image-only passed                       & $\phantom{0}2{,}156$  & $10.5\%$\\
\textbf{Image-only caught, text-only passed (novelty axis)} & $\phantom{0}\mathbf{1{,}329}$  & $\mathbf{6.5}\%$\\
Neither caught (genuinely multimodal)                     & $\phantom{0}1{,}700$  & $\phantom{0}8.3\%$\\
\bottomrule
\end{tabular}
\end{table*}

After the subsequent text-target filter restricts training to OpenI rows
with non-empty findings text, the per-row weight distribution within
the surviving $10{,}839$ rows is $95.6\%$ at $w_{\min} = 0.2$ and
$4.4\%$ at $w = 1.0$. The continuous-SFC ablation replaces the discrete
$\{w_{\min}, 1\}$ rule with $\mathrm{SFC}(x) = 1 - p_{\mathrm{text}}^{\mathrm{correct}}(x)$
clipped to $[w_{\min}, 1]$.

\subsection{IMG\_correct derivation and the ``IMG\_correct $>$ IMG'' identity}
\label{app:imgcorrect}

For a binary classifier $f$ with gold class $y$, partition examples into
$\mathrm{cc}/\mathrm{ci}/\mathrm{ic}/\mathrm{ii}$ by whether $f$ is
correct on the full and image-zeroed inputs. Then
$\mathrm{IMG} = (n_{\mathrm{ci}} - n_{\mathrm{ic}})/n \times 100$.
IMG\_correct further partitions $\mathrm{ci}$ into a margin-passing
subset (where $p_{\mathrm{full}}(y) - p_{\mathrm{masked}}(y) \geq \tau$)
and a margin-failing subset (the \emph{induced-inversion} sub-
population):
\begin{align*}
\mathrm{IMG\_correct} &= n_{\mathrm{ci,with\text{-}margin}}/n \times 100,\\
\mathrm{induced\ inv.} &= (n_{\mathrm{ci,no\text{-}margin}} -
                            n_{\mathrm{ci,with\text{-}margin}})/n \times 100.
\end{align*}
\textbf{IMG\_correct can exceed IMG.} When the $\mathrm{ic}$ population
(model becomes correct under masking) is non-trivial and the margin-
failing $\mathrm{ci}$ population is small, IMG\_correct
$> (n_{\mathrm{ci}} - n_{\mathrm{ic}})/n$. This is why several
configurations in the retroactive audit show
IMG\_correct $>$ IMG; it is mathematically valid and does \emph{not}
indicate negative induced inversion. For autoregressive generators we
apply this logic per token on the gold target sequence (with the
causal-LM shift between logits and target IDs).

\subsection{Training hyperparameters and compute}
\label{app:training}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Training hyperparameters across all five
configurations.} The configurations differ only in (i) which weights
JSONL is loaded, (ii) whether the faith loss is enabled, (iii) whether
the flipped counterfactual is enabled, and (iv) the resulting
$\lambda_{\mathrm{faith}}$. All other hyperparameters are identical.}
\label{tab:hyper}
\begin{tabular}{lc}
\toprule
\textbf{Hyperparameter} & \textbf{Value}\\
\midrule
Backbone                          & MedGemma-4B-IT (research-only licence)\\
LoRA rank / alpha                 & $16$ / $32$\\
LoRA targets                      & q\_proj, k\_proj, v\_proj, o\_proj\\
Trainable parameters              & $11.9$\,M ($0.28\%$ of $4.31$\,B base)\\
Optimizer                         & AdamW\\
Learning rate                     & $2\times 10^{-4}$\\
Weight decay                      & $0.0$ on LoRA targets, $0.01$ elsewhere\\
LR schedule                       & cosine to zero, $50$-step linear warm-up\\
Batch size                        & $2$ (grad accum $\times 8$; effective $16$)\\
Max optimizer steps               & $2{,}500$\\
Mixed precision                   & bf16\\
Faith-loss margin (nats)          & $0.3$\\
Image masking                     & per-channel mean of batch pixel values\\
$\lambda_{\mathrm{faith}}$        & $0.3$ when enabled, $0$ otherwise\\
$w_{\min}$ (filter downweight)    & $0.2$\\
Filter activation threshold       & $0.7$ on $p_{\text{branch}}^{\text{correct}}$\\
Random seeds                      & $\{42, 1337, 9001\}$\\
GPU                               & NVIDIA H100 80\,GB (Modal)\\
Wallclock per fine-tune           & $1.5$--$4.0$\,h (faith-loss configs $\sim 3\times$ slower)\\
\bottomrule
\end{tabular}
\end{table*}

Faith-loss configurations are $\sim 3\times$ slower per step than
simple-loss configurations because each step runs three forward passes
(full, image-zeroed, image-flipped). Total Phase 4 training compute:
$\approx \$200$ on Modal H100 80\,GB across $15$ fine-tunes ($5$
configs $\times$ $3$ seeds). Total project envelope (training, primary
evals, retroactive audit, cross-counterfactual, cross-site):
$\approx \$400$.

\section{Full IMG\_correct Retroactive Audit}
\label{app:audit}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Full IMG\_correct retroactive audit, all 18 cells}
(nine verifier configurations $+$ nine v5 ablations, each on $n=2{,}974$
verifier rows). \texttt{Acc full} and \texttt{Acc masked} are accuracies
under the full and image-zeroed inputs, respectively.}
\label{tab:audit-full}
\setlength{\tabcolsep}{4.0pt}
\begin{tabular}{lrrrrrr}
\toprule
\textbf{Config} & \textbf{IMG\_corr} & \textbf{IMG} & \textbf{Inv} & \textbf{Inv/IMG} & \textbf{Acc full} & \textbf{Acc mask}\\
\midrule
\multicolumn{7}{l}{\emph{Headline configurations.}}\\
v5\_0\_base                   & 3.83  & 1.28  & 0.17 & 13.2\% & 92.30 & 91.02\\
v5\_1\_ground                 & 3.40  & 2.62  & 0.13 & \phantom{0}5.1\% & 91.22 & 88.60\\
v5\_2\_real                   & 64.66 & 61.84 & 0.91 & \phantom{0}1.5\% & 92.60 & 30.77\\
\textbf{v5\_3\_contrast}      & \textbf{71.05} & \textbf{68.12} & \textbf{0.74} & \textbf{1.1\%} & 93.01 & 24.88\\
v5\_4\_final                  & 68.43 & 63.55 & 0.40 & \phantom{0}0.6\% & 91.69 & 28.14\\
v6\_0\_3site                  & 74.51 & 70.21 & 0.07 & \phantom{0}0.1\% & 91.06 & 20.85\\
v6\_0\_loo\_no\_openi         & 38.20 & 34.20 & 0.00 & \phantom{0}0.0\% & 87.42 & 53.23\\
v6\_0\_loo\_no\_padchest      & 43.95 & 41.80 & 1.31 & \phantom{0}3.1\% & 92.30 & 50.50\\
v6\_0\_loo\_no\_chestx        & 18.59 & 14.53 & 0.20 & \phantom{0}1.4\% & 81.27 & 66.75\\
\addlinespace
\multicolumn{7}{l}{\emph{v5 ablations.}}\\
abl\_scale\_25                & 69.07 & 65.64 & 0.71 & \phantom{0}1.1\% & 92.70 & 27.07\\
abl\_scale\_50                & 70.51 & 67.59 & 0.81 & \phantom{0}1.2\% & 92.80 & 25.22\\
abl\_scale\_100               & 66.24 & 62.81 & 0.61 & \phantom{0}1.0\% & 92.80 & 29.99\\
abl\_hothresh\_70             & 73.84 & 69.97 & 0.57 & \phantom{0}0.8\% & 92.60 & 22.63\\
\textbf{abl\_hothresh\_80}    & \textbf{77.98} & \textbf{74.78} & 0.74 & \phantom{0}1.0\% & 92.77 & 17.99\\
abl\_no\_contrast             & 63.28 & 60.02 & 0.87 & \phantom{0}1.4\% & 92.77 & 32.75\\
abl\_no\_ground               & 72.53 & 69.64 & 0.50 & \phantom{0}0.7\% & 92.97 & 23.34\\
abl\_no\_uncert               & 73.40 & 70.54 & 0.98 & \phantom{0}1.4\% & 93.17 & 22.63\\
\textbf{abl\_no\_consist}     & \textbf{\phantom{0}3.60}  & \textbf{\phantom{0}2.79}  & 0.30 & \phantom{0}10.8\% & 88.80 & 86.01\\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Distribution of induced-inversion contributions.} Across all
$18$ cells: induced inversion has an absolute range of
$[0.00, 1.31]$\,pp and a relative range of $[0.0\%, 13.2\%]$ of the
corresponding IMG. The cell with the largest relative induced inversion
is \texttt{v5\_0\_base} at $13.2\%$, with the smallest absolute IMG
($1.28$\,pp); fractionally large because any flipping at all under
masking is fractionally large at this scale. The cell with the largest
absolute induced inversion is \texttt{v6\_0\_loo\_no\_padchest} at
$1.31$\,pp ($3.1\%$ of IMG). No cell exceeds the pre-registered
$\geq 30\%$ threshold.

\textbf{The \texttt{abl\_no\_consist} cell is the cleanest
disconfirmation of the pre-registered concern.} Removing the consistency
loss collapses IMG\_correct from $\sim 70$\,pp (nominal v5 family) to
$3.60$\,pp --- near baseline --- while \emph{simultaneously} reducing
induced inversion from $\sim 0.7$\,pp to $0.30$\,pp. The pre-registered
concern was that consistency-loss training \emph{caused} induced
inversion. The empirical pattern is the opposite: with the consistency
loss, IMG\_correct is high and induced inversion is small; without it,
both are small.

\section{Cross-Counterfactual Detail}
\label{app:cross-cf}

\begin{table*}[t]
\centering
\small
\caption{\textbf{Cross-counterfactual eval --- full numbers including
masked accuracy.} Each row uses the same \texttt{sft\_full} seed-42
checkpoint; only the counterfactual transformation differs.}
\label{tab:cross-cf-full}
\setlength{\tabcolsep}{3.0pt}
\begin{tabular}{lcccc}
\toprule
\textbf{Counterfactual} & \textbf{IMG} & \textbf{IMG\_corr} & \textbf{Acc full} & \textbf{Acc mask}\\
\midrule
\textbf{mean (TRAINED)} & 73.97 & 74.00 & 75.48 & \phantom{0}1.51\\
noise                   & 62.10 & 62.20 & 75.48 & 13.38\\
crop                    & \phantom{0}3.64 & \phantom{0}4.81 & 75.48 & 71.83\\
occlude                 & \phantom{0}0.28 & \phantom{0}1.11 & 75.48 & 75.20\\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Baseline anchor.} On the matching crop counterfactual,
\texttt{sft\_only} seed-42 yields IMG $= 0.33$\,pp, IMG\_correct
$= 2.80$\,pp at $73.82\%$ accuracy, induced inversion $= 0.26$\,pp
($n = 105{,}921$ tokens). The cross-counterfactual \texttt{sft\_full}
$-$ \texttt{sft\_only} IMG\_correct gap on crop is
$4.81 - 2.80 = 2.01$\,pp; the gap on the trained-mean counterfactual is
$74.00 - 2.38 = 71.62$\,pp. The order-of-magnitude difference between
trained-mean and crop is the operational meaning of the partial-
circularity finding.

\section{Cross-Site Detail}
\label{app:crosssite}

\textbf{Silver target construction.} ChestX-Det10 test rows in the
standard test set have empty \texttt{evidence\_text} fields. We
constructed \texttt{groundbench\_test\_chestx\_det10\_silver.jsonl}
($1{,}122$ rows) by injecting CheXagent-2-3B
\citep{chen2024chexagent} generations as \texttt{reference\_report};
the silver-target nature is reported explicitly. PadChest-GR cross-site
\citep{castro2024padchestgr} requires a fresh CheXagent generation pass
and is deferred. \textbf{Token-level analysis.} The cross-site $54.65\%$
token-level accuracy is $20$\,pp lower than the in-distribution OpenI
accuracy, reflecting (i) genuine cross-site difficulty (different
camera vendors, different radiologist phrasing) and (ii) the silver-
target nature (the model is being graded against another model's
output, not radiologist gold).

\section{Pre-Registered Hypotheses}
\label{app:prereg}

\textbf{(i) Full method reduces IMG\_correct on OpenI-test by
$\geq 5$\,pp vs.\ SFT-only at matched accuracy.} Status:
\textbf{HIT BY AN ORDER OF MAGNITUDE} on three seeds $\times$ three
faith-loss configurations. Realised: $+72.58$\,pp at $+1.17$\,pp
accuracy.

\textbf{(ii) IMG\_correct refinement reveals $\geq 30\%$ of
\texttt{v5\_3\_contrast}'s IMG as induced inversion.} Status:
\textbf{REFUTED}. Realised: $1.1\%$. Pattern holds across all $18$
audited cells (induced-inversion contributions $\leq 1.31$\,pp
absolute, $\leq 13.2\%$ relative).

\textbf{(iii) Dual filter dominates the faith loss in the ablation.}
Status: \textbf{INVERTED}. The empirical pattern is the opposite:
\texttt{sft\_dual} is statistically indistinguishable from
\texttt{sft\_only}; the faith loss is the load-bearing component
(\texttt{sft\_faith} alone gives $+71.73$\,pp over baseline). The
increment from adding the dual filter on top of the faith loss is
$+0.85$\,pp seed-mean (within one standard deviation of
\texttt{sft\_full}). We retain the dual filter in the recipe because
the four-cell decomposition (App.~\ref{app:filter}) demonstrates a
symmetric-extension contribution axis not addressable by text-only
filtering.

One pre-registration HIT, one REFUTED, one INVERTED --- all reported
transparently. The methodological refinement (IMG\_correct) lands as a
contribution independently of the empirical results.

\bibliographystyle{icml2026}
\bibliography{pivot_ab_references}

\end{document}
